Conformal Counterfactual Explanations

### 2023-03-25 paper/paper.tex
uh

This is typically implemented by defining a target outcome $t \in \mathcal{Y}$ for some individual $x \in \mathcal{X}$, for which the model $M_{\theta}:\mathcal{X}\mapsto\mathcal{Y}$ initially predicts a different outcome: $M_{\theta}(x)\ne t$. Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(x),t)$. Since Counterfactual Explanations (CE) work directly with the Black Box Model, they always have full local fidelity by construction. Fidelity is defined as the degree to which explanations approximate the predictions of the Black Box Model. This arguably one of the most important evaluation metrics for model explanations, since any explanation that explains a prediction not actually made by the model is useless \citep{molnar2020interpretable}.

@@

@@

\begin{itemize}
  \item What exact sampler do we use? ImproperSGLD as in \citet{grathwohl2020your} seems to work best.
  \item How exactly do we plan to quantify plausibility and conformity? Elaborate on measures.
\end{itemize}

@@

\subsection{Conformal Training meets Counterfactual Explanations}

Now that we have a framework for evaluating Counterfactual Explanations in terms of their plausibility and conformity, we are interested in finding a way to generate counterfactuals that are as plausible and conformal as possible. We hypothesize that a narrow focus on plausibility may come at the cost of reduced conformity. Using a surrogate model for the generative task, for example, may improve plausibility but inadvertently yield counterfactuals that are more consistent with the surrogate than the Black Box Model itself.

While we do not want seek to discourage the use of surrogate models, we suggest that one way to ensure model conformity is to rely strictly on the model itself.~\citet{schut2021generating} demonstrate that this restriction need not impede plausibility, since we can rely on predictive uncertainty estimates to guide our counterfactual search. By avoiding counterfactual paths that are associated with high predictive uncertainty, we end up generating counterfactuals for which the model $M_{\theta}$ predicts the target label $t$ with high confidence. Provided the model is well-calibrated, these counterfactuals are plausible which the authors demonstrate empirically through benchmarks.

This idea hinges on ...

>>>>

This is typically implemented by defining a target outcome $t \in \mathcal{Y}$ for some individual $x \in \mathcal{X}$, for which the model $M_{\theta}:\mathcal{X}\mapsto\mathcal{Y}$ initially predicts a different outcome: $M_{\theta}(x)\ne t$. Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(x),t)$. Since Counterfactual Explanations (CE) work directly with the Black Box Model, valid counterfactuals always have full local fidelity by construction \citep{mothilal2020explaining}. Fidelity is defined as the degree to which explanations approximate the predictions of the Black Box Model. This is arguably one of the most important evaluation metrics for model explanations, since any explanation that explains a prediction not actually made by the model is useless \citep{molnar2020interpretable}.

@@

\section{Evaluating the Faithfulness of Counterfactuals}\label{conformity}

@@

\begin{itemize}
  \item What exact sampler do we use? ImproperSGLD as in \citet{grathwohl2020your} seems to work best.
\end{itemize}

@@

Now that we have a framework for evaluating Counterfactual Explanations in terms of their plausibility and conformity, we are interested in finding a way to generate counterfactuals that are as plausible and conformal as possible. We hypothesize that a narrow focus on plausibility may come at the cost of reduced conformity. Using a surrogate model for the generative task, for example, may improve plausibility but inadvertently yield counterfactuals that are more consistent with the surrogate than the Black Box Model itself. We suggest that one way to ensure model conformity is to rely strictly on the model itself. In this section, we introduce a novel framework that meets this requirement, works under minimal assumptions and does not impede the plausibility objective: Conformal Counterfactual Explanations.

\subsection{Predictive Uncertainty Quantification for Counterfactual Explanations}

Our proposed methodolgy is built off of the findings presented in ~\citet{schut2021generating}. The authors demonstrate that it is not only possible but remarkably easy to generate plausible counterfactuals for Black Box Models that provide predictive uncertainty estimates. By avoiding counterfactual paths that are associated with high predictive uncertainty, we end up generating counterfactuals for which the model $M_{\theta}$ predicts the target label $t$ with high confidence. Provided the model is well-calibrated, these counterfactuals are plausible which the authors demonstrate empirically through benchmarks.

\textbf{TBD}: maximizing predicted probability is equivalent to minimizing uncertainty ...

The approach proposed by \citet{schut2021generating} hinges on the crucial assumption that the Black Box Model provides predictive uncertainty estimates. The authors argue that in light of rapid advances in Bayesian Deep Learning (DL), this assumption is overall less costly than the engineering overhead induced by using surrogate models. This is even more true today, as recent work has put Laplace Approximation back on the map for truly effortless Bayesian DL \citep{immer2020improving,daxberger2021laplace,antoran2023sampling}. Nonetheless, the need for Bayesian methods may be too restrictive in some cases.

Fortunately, there is a promising alternative approach to predictive uncertainty quantification (UQ) that we will turn to next: Conformal Prediction.

\subsection{Conformal Prediction}

Conformal Prediction (CP): a scalable and statistically rigorous  Not only is CP statistically rigorous and works under minimal distributional assumptions, it is also model-agnostic and can be applied at test time. That last part is critical since it allows us to relax the assumption that the Black Box Model needs to learn to generate predictive uncertainty estimates during training. In other words, CP promises to provide a way to generate plausible counterfactuals for any model without the need for surrogate models.

\subsection{Conformal Counterfactual Explanations}
